---
title: Syllabus
slug: syllabus
citeproc: true
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    toc: true
    fig_caption: false
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(tibble)
library(pander)
```

`r blogdown::shortcode("courseinfo")`

## Course objectives

The public sector is complex. Public administrators, managers, and policy makers need to be fluent in the language of economics and need to be able to engage in and understand quantitative analysis of social policies. 

In this class, you'll learn how to *understand*, *speak*, and *do* economics.

By the end of this course, you will (1) be literate in fundamental economic principles, (2) understand the limits of economic theory and free markets, (3) justify government and nonprofit intervention in the economy, and (4) make informed policy recommendations by analyzing and evaluating public sector policies. Specifically, you'll be able to:

- **Understand** the principles of microeconomics, public economics, and behavioral economics
- **Explain** social phenomena using economic vocabulary and reasoning
- **Predict** how individuals respond to incentives
- **Justify** government intervention in the free market and **identify** when public policies have been unethical or failures
- **Propose and argue** for public and nonprofit sector policies

## Course materials

Most of the readings in this class are **free**.

### Books

We will only use one physical textbook. There are two official textbooks for the class:

- [@Wheelan:2019]

    > <span>$11</span> used, <span>$12</span> new at [Amazon](https://www.amazon.com/Naked-Economics-Undressing-Dismal-Science/dp/0393356493/). The 2019 edition is preferred; the 2010 edition should be okay too, since the chapters line up.

- [@core-espp]

    > [**FREE** online](https://www.core-econ.org/espp/).

[CORE Econ](https://www.core-econ.org/) is a special new project that aims to make economics education accessible to all, replacing textbooks that cost hundreds of dollars with an open source textbook complete with videos and quizzes and other online resources. It's even been [lauded by The Economist](https://www.economist.com/news/finance-and-economics/21729443-students-are-forced-grapple-real-world-very-beginning-their).

CORE's original book, [*The Economy*](http://www.core-econ.org/the-economy/), was designed to serve as a 1–2 semester introduction to economics for economics majors. CORE recently created a version of their materials *specifically* for those interested in public administration and policy. [*Economy, Society, and Public Policy*](https://www.core-econ.org/espp/) is designed for non-economics majors who have no interest in becoming economists, but who want to understand economics and policy. This is an ideal book for our class and I'm so excited to use it. It's still a beta project, and there might be errors and quirks and bumps in the road, but (1) it's free, and (2) it's state of the art and you're some of the first students to ever use it. So live with the quirks :)

[**Register for a student account at the CORE website.**](http://www.core-econ.org/register/)

### Articles and other materials

There will occasionally be additional articles and videos to read and watch. When this happens, links to these other resources will be included on the content page for that session.


## Course policies

**Be nice. Be honest. Don't cheat.**

We will also follow [Georgia State's Code of Conduct](https://codeofconduct.gsu.edu/).

This syllabus reflects a plan for the semester. Deviations may become necessary as the course progresses.

### Student hours

Please watch this video:

<iframe src="https://player.vimeo.com/video/270014784" width="640" height="360" frameborder="0" allow="autoplay; fullscreen" allowfullscreen style="display: block; margin: 0 auto 1rem;"></iframe>

Student hours are set times dedicated to all of you (most professors call these "office hours"; I don't^[There's fairly widespread misunderstanding about what office hours actually are! [Many students often think that they are the times I *shouldn't* be disturbed](https://www.chronicle.com/article/Can-This-Man-Change-How-Elite/245714/), which is the exact opposite of what they're for!]). This means that I will be ~~in my office~~ at home (wistfully) waiting for you to ~~come by~~ talk to me remotely with whatever questions you have. This is the best and easiest way to find me and the best chance for discussing class material and concerns.

Because of the pandemic, we cannot meet in person. I can meet you online via Webex. [Make an appointment with me here](https://calendly.com/andrewheiss/), and then use this link to talk to me during student hours: <https://gsumeetings.webex.com/meet/aheiss>. You can also find me through e-mail and Slack. 


### Learning during a pandemic

Life absolutely sucks right now. None of us is really okay. **We're all just pretending.**

You most likely know people who have lost their jobs, have tested positive for COVID-19, have been hospitalized, or perhaps have even died. You all have increased (or possibly decreased) work responsibilities and increased family care responsibilities—you might be caring for extra people (young and/or old!) right now, and you are likely facing uncertain job prospects (or have been laid off!). 

**I'm fully committed to making sure that you learn everything you were hoping to learn from this class!** I will make whatever accommodations I can to help you finish your exercises, do well on your projects, and learn and understand the class material. Under ordinary conditions, I am flexible and lenient with grading and course expectations when students face difficult challenges. Under pandemic conditions, that flexibility and leniency is intensified.

If you tell me you're having trouble, I will not judge you or think less of you. I hope you'll extend me the same grace. 

You *never* owe me personal information about your health (mental or physical). You are *always* welcome to talk to me about things that you're going through, though. If I can't help you, I usually know somebody who can.

If you need extra help, or if you need more time with something, or if you feel like you're behind or not understanding everything, **do not suffer in silence!** Talk to me! I will work with you. **I promise.** 

*Please* sign up for a time to meet with me during student hours at <https://calendly.com/andrewheiss/meeting/>. I'm also available through e-mail and Slack. I've enabled notifications on my Slack account, so I'll see your messages quickly!

I want you to learn lots of things from this class, but I primarily want you to stay healthy, balanced, and grounded during this crisis.

### Late work

You will lose 1 point per day for each day a problem set is late (that's designed to not be a huge penalty—3 days late = 17/20 points on a problem set that gets a ✓).

After 1 week, I will send an automated reminder e-mail. After 2 weeks, you will receive no points.


### Counseling and Psychological Services (CPS)

Life at GSU can be complicated and challenging (especially during a pandemic!). You might feel overwhelmed, experience anxiety or depression, or struggle with relationships or family responsibilities. [Counseling and Psychological Services (CPS)](https://education.gsu.edu/cps/) provides free, *confidential* support for students who are struggling with mental health and emotional challenges. The CPS office is staffed by professional psychologists who are attuned to the needs of all types of college and professional students. Please do not hesitate to contact CPS for assistance—getting help is a smart and courageous thing to do.

### Basic needs security

If you have difficulty affording groceries or accessing sufficient food to eat every day, or if you lack a safe and stable place to live, and you believe this may affect your performance in this course, please contact the [Dean of Students](https://deanofstudents.gsu.edu/) for support. They can provide a host of services including free groceries from the [Panther Pantry](https://nutrition.gsu.edu/panther-pantry/) and assisting with homelessness with the [Embark Network](https://deanofstudents.gsu.edu/student-assistance/embark/). Additionally, please talk to me if you are comfortable in doing so. This will enable me to provide any resources that I might possess.

### Lauren's Promise

**I will listen and believe you if someone is threatening you.**

Lauren McCluskey, a 21-year-old honors student athlete, [was murdered on October 22, 2018 by a man she briefly dated on the University of Utah campus](https://www.sltrib.com/opinion/commentary/2019/02/10/commentary-failing-lauren/). We must all take action to ensure that this never happens again.

If you are in immediate danger, call 911 or GSU police (404-413-3333).

If you are experiencing sexual assault, domestic violence, or stalking, please report it to me and I will connect you to resources or call [GSU's Counseling and Psychological Services](https://counselingcenter.gsu.edu/crisis-services/concern-self/immediate-help/) (404-413-1640).

Any form of sexual harassment or violence will not be excused or tolerated at Georgia State. GSU has instituted procedures to respond to violations of these laws and standards, programs aimed at the prevention of such conduct, and intervention on behalf of the victims. Georgia State University Police officers will treat victims of sexual assault, domestic violence, and stalking with respect and dignity. Advocates on campus and in the community can help with victims' physical and emotional health, reporting options, and academic concerns.

### Academic honesty

Violation of [GSU's Policy on Academic Honesty](https://deanofstudents.gsu.edu/faculty-staff-resources/academic-honesty/) will result in an F in the course and possible disciplinary action.^[So seriously, just don't cheat or plagiarize!] All violations will be formally reported to the Dean of Students.

### Special needs

Students who wish to request accommodation for a disability may do so by registering with the [Office of Disability Services](https://disability.gsu.edu/). Students may only be accommodated upon issuance by the Office of Disability Services of a signed [Accommodation Plan](https://disability.gsu.edu/services/how-to-register/) and are responsible for providing a copy of that plan to instructors of all classes in which accommodations are sought.

Students with special needs should then make an appointment with me during the first week of class to discuss any accommodations that need to be made.


## Assignments and grades

You can find descriptions for all the assignments on the [assignments page](/assignment/).

```{r assignments-grades, include=FALSE}
assignments <- tribble(
  ~Points,  ~Assignment,
  10 * 14  , "Weekly reports (10 × 14)",
  20 * 7  , "Problem sets (20 × 7)",
  150     , "Exam 1",
  150     , "Exam 2"
) %>%
  mutate(Percent = Points / sum(Points),
         Percent = scales::percent(Percent, accuracy = 1))

grading <- tribble(
  ~Grade, ~Range,  ~Grade1, ~Range1,
  "A",  "93–100%", "C",  "73–76%",
  "A−", "90–92%",  "C−", "70–72%",
  "B+", "87–89%",  "D+", "67–69%",
  "B",  "83–86%",  "D",  "63–66%",
  "B−", "80–82%",  "D−", "60–62%",
  "C+", "77–79%",  "F",  "< 60%"
)
```

<div class="centered-table">

```{r show-assignments-table, echo=FALSE, results="asis"}
assignments_total <- tibble(Assignment = "Total",
                            Points = sum(assignments$Points),
                            Percent = NA)

bind_rows(assignments, assignments_total) %>% 
  select(Assignment, Points, Percent) %>% 
  pandoc.table(justify = "lcc", missing = "—")
```
</div>

<div class="centered-table">

```{r show-grades-table, echo=FALSE, results="asis"}
grading %>%
  pandoc.table(justify = "lclc", 
               col.names = c("Grade", "Range", "Grade", "Range"))
```

<div>


## Movies

Once you have read this entire syllabus and [the assignment overview page](/assignment/), please [click here](mailto:aheiss@gsu.edu) and e-mail me the poster for one of your favorite movies.
